---
title: "SIG04 Seurat from anndata"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(hdf5r)
library(ggplot2)
library(clustree)
library(cowplot)
library(patchwork)
library(ggpointdensity)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
source("functions/scRNA_seq_analysis_functions.R")
theme_set(theme_Publication())
```

Here, I will set up a seurat object with just counts (no SCT transform) and the metadata and cells from the anndata object. 

## [Import data]{.underline}

```{r}
# import raw counts from cellranger
dataRaw1 <- Read10X_h5("C:/Users/Eric/Documents/datasets/EYW/SIG04_10x_240816/cellranger_outs/SIG04_lane1_merge_GEX_oBC/filtered_feature_bc_matrix.h5")
dataRaw2 <- Read10X_h5("C:/Users/Eric/Documents/datasets/EYW/SIG04_10x_240816/cellranger_outs/SIG04_lane2_merge_GEX_oBC/filtered_feature_bc_matrix.h5")

# import post-QC anndata metadata
metaLane1 <- read.csv("processing_outs/postQC_meta_lane1.csv", row.names = "cell_barcode")
metaLane2 <- read.csv("processing_outs/postQC_meta_lane2.csv", row.names = "cell_barcode")

names(dataRaw1)
names(dataRaw2)
```

## [Create Seurat]{.underline}

### Filter for joint cells

```{r}
subset_joint <- function(x){
  # identify joint cells between BC and GEX dataset
  jointCells <- intersect(colnames(x$`Gene Expression`),
                          colnames(x$`CRISPR Guide Capture`))
  
  # subset each dataset to include joint cells
  data <- vector(mode = "list")
  data[["GEX"]] <- x$`Gene Expression`[,jointCells]
  data[["oBC"]] <- x$`CRISPR Guide Capture`[,jointCells]
  
  return(data)
}

dataRaw1 <- subset_joint(dataRaw1)
dataRaw2 <- subset_joint(dataRaw2)
```

### Filter Low Genes

Remove genes that are expressed by less than 1 in 1000 cells. Set this level based on the smallest size cluster that you want to be
detectable in the final dataset.

```{r}
dataMergeGEX <- cbind(dataRaw1$GEX,dataRaw2$GEX)

# create vector of genes to keep
keepGenes <- Matrix::rowSums(dataMergeGEX > 0) >= ncol(dataMergeGEX)*(1/1000)

# remove genes from gex matrices
dataRaw1$GEX <- dataRaw1$GEX[keepGenes,]
dataRaw2$GEX <- dataRaw2$GEX[keepGenes,]

rm(dataMergeGEX)
```

### Create Seurat Object

```{r}
create_seurat <- function(x){
  # load GEX counts and add celltag counts into seurat object
  data <- CreateSeuratObject(counts = x$GEX, project = "SIG04")
  data[["oBC"]] <- CreateAssayObject(counts = x$oBC)

  return(data)
}

dataSeurat <- vector(mode = "list")
dataSeurat[["lane1"]] <- create_seurat(dataRaw1)
dataSeurat[["lane2"]] <- create_seurat(dataRaw2)

rm(dataRaw1,dataRaw2)
```

### Filter and merge with Anndata Metadata

```{r}
merge_adMeta <- function(seurat,meta){
  seurat <- subset(seurat, cells = rownames(meta))
  # make sure cell barcodes match order between metadata and seurat
  meta <- meta[match(colnames(seurat),rownames(meta)),]
  print(paste0("All cells match?: ",
               sum(rownames(meta) == colnames(seurat)) == ncol(seurat)))
  # replace metadata with anndata meta
  seurat@meta.data <- meta
  
  return(seurat)
}

dataSeurat[["lane1"]] <- merge_adMeta(dataSeurat$lane1, metaLane1)
dataSeurat[["lane2"]] <- merge_adMeta(dataSeurat$lane2, metaLane2)
```

### Merge 10x lanes into single dataset

```{r}
data <- merge(dataSeurat[[1]],dataSeurat[[2]],
              add.cell.ids = c("lane1","lane2"), project = "SIG04") %>%
  JoinLayers()

rm(dataSeurat)
```

## [Scale and Normalize data]{.underline}

```{r}
DefaultAssay(data) <- "RNA"
data <- NormalizeData(data, normalization.method = "LogNormalize") %>%
  FindVariableFeatures(selection.method = "vst",
                       nfeatures = 2000, 
                       verbose = FALSE) %>%
  ScaleData
data <- ScaleData(data)

DefaultAssay(data) <- "oBC"
data <- NormalizeData(data, normalization.method = "CLR") %>%
  ScaleData
```

## [Score for sources of variation]{.underline}

### Cell cycle variation
```{r}
DefaultAssay(data) <- "RNA"

cellCycleGenes <- read_csv("C:/Users/Eric/Documents/datasets/gene_sets/cell_cycle_genes_seurat_mouse.csv")

data <- CellCycleScoring(data, s.features = cellCycleGenes$s_genes,
                            g2m.features = cellCycleGenes$g2m_genes)
```

```{r}
saveRDS(data, "C:/Users/Eric/Documents/datasets/EYW/SIG04_10x_240816/seurat_outs/SIG04_postQC_seurat_all.rds")
```
