---
title: "SIG04 10x Seurat Processing"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(hdf5r)
library(ggplot2)
library(clustree)
library(cowplot)
library(patchwork)
library(ggpointdensity)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
source("functions/scRNA_seq_analysis_functions.R")
theme_set(theme_Publication())
```

## [Import data]{.underline}

```{r}
dataRaw1 <- Read10X_h5("C:/Users/Eric/My Drive/Lab/datasets/EYW/SIG04_10x_240816/cellranger_outs/SIG04_lane1_GEX_oBC/filtered_feature_bc_matrix.h5")
dataRaw2 <- Read10X_h5("C:/Users/Eric/My Drive/Lab/datasets/EYW/SIG04_10x_240816/cellranger_outs/SIG04_lane2_GEX_oBC/filtered_feature_bc_matrix.h5")

names(dataRaw1)
names(dataRaw2)
```

```{r}
subset_joint <- function(x){
  # identify joint cells between BC and GEX dataset
  jointCells <- intersect(colnames(x$`Gene Expression`),
                          colnames(x$`Antibody Capture`))
  
  # subset each dataset to include joint cells
  data <- vector(mode = "list")
  data[["GEX"]] <- x$`Gene Expression`[,jointCells]
  data[["oBC"]] <- x$`Antibody Capture`[,jointCells]
  
  return(data)
}

dataRaw1 <- subset_joint(dataRaw1)
dataRaw2 <- subset_joint(dataRaw2)
```


### Filter Low Genes

Remove genes that are expressed by less than 1 in 5000 cells (assuming 30% dropout
rate). Set this level based on the smallest size cluster that you want to be
detectable in the final dataset.

```{r}
dataMergeGEX <- cbind(dataRaw1$GEX,dataRaw2$GEX)

# create vector of genes to keep
keepGenes <- Matrix::rowSums(dataMergeGEX > 0) >= ncol(dataMergeGEX)*0.0002*0.3

# remove genes from gex matrices
dataRaw1$GEX <- dataRaw1$GEX[keepGenes,]
dataRaw2$GEX <- dataRaw2$GEX[keepGenes,]

rm(dataMergeGEX)
```

### Filter Low oBC

```{r}
sum(colSums(dataRaw1$oBC > 0) > 0)/ncol(dataRaw1$oBC)
sum(colSums(dataRaw2$oBC > 0) > 0)/ncol(dataRaw2$oBC)
```


```{r}
create_seurat <- function(x){
  # load GEX counts and add celltag counts into seurat object
  data <- CreateSeuratObject(counts = x$GEX, project = "SIG04")
  data[["oBC"]] <- CreateAssayObject(counts = x$oBC)

  return(data)
}

dataSeurat <- vector(mode = "list")
dataSeurat[["lane1"]] <- create_seurat(dataRaw1)
dataSeurat[["lane2"]] <- create_seurat(dataRaw2)

rm(dataRaw1,dataRaw2)
```

## [Cell QC Analysis]{.underline}

### Calculate QC metrics

```{r}
add_qc_metrics <- function(x){
  DefaultAssay(x) <- "RNA"
  # calculate percent mitochondrial reads and ratio 
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^mt-")
  x[["percent.ribo"]] <- PercentageFeatureSet(x, pattern = "^Rpl|^Rps")
  x[["mitoRatio"]] <- x@meta.data$percent.mt / 100
  x[["log10GenesPerUMI"]] <- log10(x$nFeature_RNA)/log10(x$nCount_RNA)
  x[["log10nFeature_RNA"]] <- log10(x$nFeature_RNA)
  x[["log10nCount_RNA"]] <- log10(x$nCount_RNA)
  
  return(x)
}

dataSeurat <- lapply(dataSeurat, function(x) add_qc_metrics(x))

metadata <- bind_rows(dataSeurat[[1]]@meta.data,dataSeurat[[2]]@meta.data) %>%
  mutate(orig.ident = c(rep("lane1",ncol(dataSeurat[[1]])),rep("lane2",ncol(dataSeurat[[2]]))))
```

### Visualize QC distributions

```{r, fig.width=15, fig.height=7}
# enter QC values
qcValues <- c(nFeature_RNA = 1000, nCount_RNA = 2000, percent.mt = 5)

# quantify how many cells there are
paste0("There are ", ncol(data), " cells")

# Visualize the number UMIs/transcripts per cell
p1 <- metadata %>%
  	ggplot(aes(color=orig.ident, x=nCount_RNA, fill= orig.ident)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	ylab("Cell density") +
  	geom_vline(xintercept = qcValues[2]) +
    NoLegend()

# Visualize the distribution of genes detected per cell via histogram
p2 <- metadata %>% 
  	ggplot(aes(color=orig.ident, x=nFeature_RNA, fill= orig.ident)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	geom_vline(xintercept = qcValues[1]) +
    NoLegend()

# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
p5 <- metadata %>% 
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  	geom_point() + 
  	scale_colour_gradient(low = "gray90", high = "black") +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	geom_vline(xintercept = qcValues[2]) +
  	geom_hline(yintercept = qcValues[1])


# Visualize the distribution of mitochondrial gene expression detected per cell
p3 <- metadata %>% 
  	ggplot(aes(color=orig.ident, x=percent.mt, fill=orig.ident)) + 
  	geom_density(alpha = 0.2) + 
  	geom_vline(xintercept = qcValues[3]) +
    xlim(0,10) +
    NoLegend()

# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
p4 <- metadata %>%
  	ggplot(aes(x=log10GenesPerUMI, color = orig.ident, fill=orig.ident)) +
  	geom_density(alpha = 0.2) +
  	geom_vline(xintercept = 0.8)

# Visualize the correlation between genes detected and number of UMIs with mt filter
p6 <- metadata %>%
    filter(percent.mt < qcValues[3]) %>%
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA)) + 
    geom_pointdensity() +
    scale_color_viridis_c() +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	geom_vline(xintercept = qcValues[2]) +
  	geom_hline(yintercept = qcValues[1])

plot_grid(p1,p2,p3,p4,p5,p6, ncol = 3)
```

```{r}
metadata %>%
  	ggplot(aes(color=orig.ident, x=nCount_oBC, fill= orig.ident)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	ylab("Cell density")
```
It's weird that lane 2 which consists of sorted cells has such a skewed distribution compared to lane 1. Theoretically you would think that sorting would improve barcode detection. 

### Subset data based on QC metrics

```{r}
dataQC <- lapply(dataSeurat, function(x) subset(x, subset = (nFeature_RNA >= qcValues[1]) & (nCount_RNA >=qcValues[2]) & (percent.mt < qcValues[3])))
```

Remove cells that have 0 oBC counts

```{r}
paste0(ncol(subset(dataQC[[2]], subset = nCount_oBC > 0))/ncol(dataQC[[2]])," of cells have oBC > 0")

dataQC[[2]] <- subset(dataQC[[2]], subset = nCount_oBC > 0)
```


## [HTODemux oBC]{.underline}

Use oBC pseudo because HTODemux will throw an error if a feature has no counts. oBC pseudo has an additional count added to all cells for all features

```{r}
dataQC <- lapply(dataQC, function(x) NormalizeData(x, assay = "oBC", normalization.method = "CLR"))
dataQC <- lapply(dataQC, function(x) HTODemux(x, assay = "oBC"))

test <- dataQC[[2]]
test <- HTODemux(test, assay = "oBC", nsamples = 1000)

```

```{r, fig.width=6, fig.height=5}
# Calculate a UMAP embedding of the HTO data
DefaultAssay(dataQC$lane1) <- "oBC"
DefaultAssay(dataQC$lane2) <- "oBC"

dataQC <- lapply(dataQC, function(x) ScaleData(x, features = rownames(x),
                                                       verbose = FALSE)) %>%
  lapply(function(x) RunPCA(x, features = rownames(x), approx = FALSE, verbose = F)) %>%
  lapply(function(x) RunUMAP(x, dims = 1:nrow(x), verbose = F))
```

```{r, fig.width=12, fig.height=8}
plotHTO <- function(x){
  # create heatmap of HTO expression
  p1 <- HTOHeatmap(x, assay = "HTO")
  Idents(x) <- "HTO_classification.global"
  # compare UMI number for singlets, doublets, negatives
  p2 <- VlnPlot(x, features = "nCount_RNA", pt.size = 0.1, log = TRUE, alpha = 0.5)
  # plot HTO data on tSNE
  p3 <- DimPlot(x)
  
  (p3+p2)/p1
}

plotHTO(dataSeurat$lane1)
plotHTO(dataSeurat$lane2)
```

```{r, fig.width=12, fig.height=16}
# vsiualize ridge plots for HTO
Idents(dataSeurat$lane1) <- "HTO_maxID"
Idents(dataSeurat$lane2) <- "HTO_maxID"

lapply(dataSeurat, function(x) RidgePlot(x, assay = "HTO", features = rownames(x[["HTO"]]), ncol = 3))
```




## [Merge 10x lanes into single dataset]{.underline}

```{r}
data <- merge(dataSeurat[[1]],dataSeurat[[2]],
              add.cell.ids = c("lane1","lane2"), project = "SIG04") %>%
  JoinLayers()

# rename orig.ident based on
data$orig.ident <- paste(data$orig.ident, gsub("_.*","", colnames(data)), sep = "-")

rm(dataSeurat)
```




## [Remove HTO negatives]{.underline}

```{r}
Idents(dataSub) <- "HTO_classification.global"
dataSub <- subset(dataSub, idents = "Negative", invert = T)
```

```{r}
dataSub@meta.data %>%
  group_by(hash.ID) %>%
  summarize(num_cells = n()) %>%
  mutate(percent = num_cells/sum(num_cells)*100)
```


## [Score for sources of variation]{.underline}

### Cell cycle variation
```{r}
DefaultAssay(dataSubQC) <- "RNA"
dataSubQC <- NormalizeData(dataSubQC)
# Identify the most variable genes
dataSubQC <- FindVariableFeatures(dataSubQC, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
# Scale the counts
dataSubQC <- ScaleData(dataSubQC)

cellCycleGenes <- read_csv("C:/Users/Eric/My Drive/Lab/datasets/gene_sets/cell_cycle_genes_seurat_mouse.csv")
# cellCycleGenes <- read_csv("/Users/wange7/Library/CloudStorage/GoogleDrive-ericwang314@gmail.com/My Drive/Lab/datasets/gene_sets/cell_cycle_genes_seurat_mouse.csv")

dataSubQC <- CellCycleScoring(dataSubQC, s.features = cellCycleGenes$s_genes,
                            g2m.features = cellCycleGenes$g2m_genes)
```

```{r, fig.width=8, fig.height=4}
# Plot the PCA colored by cell cycle phase
dataSubQC <- RunPCA(dataSubQC, verbose = FALSE)
DimPlot(dataSubQC,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")
```

### Mitochondrial variation

```{r, fig.width = 12, fig.height=4}
# Check quartile values
mitoQuart <- summary(dataSubQC@meta.data$mitoRatio)

# Turn mitoRatio into categorical factor vector based on quartile values
dataSubQC@meta.data$mitoFr <- cut(dataSubQC@meta.data$mitoRatio, 
                   breaks=c(-Inf, mitoQuart[2], mitoQuart[3], mitoQuart[5], Inf), 
                   labels=c("Low","Medium","Medium high", "High"))

DimPlot(dataSubQC,
        reduction = "pca",
        group.by= "mitoFr",
        split.by = "mitoFr")
```

## [Data normalization/scaling]{.underline}

### SCTransform

```{r}
options(future.globals.maxSize= 10000*1024^2)
# perform SCT normalization and find variable features
dataSubQC <- SCTransform(dataSubQC, vst.flavor = "v2",
                       vars.to.regress = c("orig.ident","mitoRatio","Phase"),
                       variable.features.n = 3100)

# remove TCR genes from variable features
featureVar <- VariableFeatures(dataSubQC, assay = "SCT")
TCRgenes <- grep("^Tra[vj]|^Trb[vdj]",featureVar)
BCRgenes <- grep("^Igk[vc]|^Igl[vc]|^Ighv",featureVar)
featureVar[TCRgenes]
featureVar[BCRgenes]

VariableFeatures(dataSubQC, assay = "SCT") <- featureVar[-c(TCRgenes,BCRgenes)]

# scale SCT variable features in RNA assay
DefaultAssay(dataSubQC) <- "RNA"
dataSubQC <- ScaleData(dataSubQC, features = featureVar[-c(TCRgenes,BCRgenes)])
DefaultAssay(dataSubQC) <- "SCT"
```

### CITE

```{r}
dataSubQC <- NormalizeData(dataSubQC, assay = "CITE", normalization.method = "CLR")
```

```{r, fig.height=6, fig.width=10}
RidgePlot(dataSubQC, assay = "CITE", features = c("CD62L","CD44"), group.by = "HTO_maxID")
```

## [Dim Reduction]{.underline}

```{r}
# run PCA using TCR/BCR excluded Variable Features
dataSubQC <- RunPCA(dataSubQC, npcs = 60, verbose = FALSE)
ElbowPlot(dataSubQC, ndims = 60) +
  ggtitle("dataSubQC Elbow") +
  scale_x_continuous(breaks = seq(0,60,3))
```

```{r}
dimValues <- 54
# run umap based on dimensionality from elbow plot
dataSubQC <- RunUMAP(dataSubQC, dims = 1:dimValues, verbose = F) %>%
  FindNeighbors(dims = 1:dimValues, verbose = FALSE) %>%
  FindClusters(resolution = seq(0,1.2,0.1),
               algorithm = 1)
```

```{r, fig.height=10, fig.width=6}
clustree(dataSubQC, prefix = "SCT_snn_res.") +
  ggtitle("dataSubQC Clustree")
```

```{r, fig.width=12, fig.height=10}
dataSubQC$seurat_clusters <- dataSubQC$SCT_snn_res.0.8
Idents(dataSubQC) <- dataSubQC$seurat_clusters
p1 <- DimPlot(dataSubQC) + theme(aspect.ratio = 1)
p2 <- DimPlot(dataSubQC, group.by = "hash.ID") + theme(aspect.ratio = 1)
p3 <- DimPlot(dataSubQC, group.by = "crispr_classification") + theme(aspect.ratio = 1)
(p1+p2)/(p3)
```

```{r, fig.height=25, fig.width=10}
# perform DEG analysis on SCT data
dataSubQC <- PrepSCTFindMarkers(dataSubQC)
clusMarkers <- FindAllMarkers(dataSubQC, only.pos = T, assay = "SCT")
clusMarkers <- as_tibble(clusMarkers)

sigMarkers <- clusMarkers %>%
  dplyr::filter(avg_log2FC > 1 & p_val_adj < 0.1) %>%
  arrange(p_val_adj)

top10 <- sigMarkers %>%
    group_by(cluster) %>%
    slice_head(n = 10) %>%
    ungroup()

top10
```

```{r}
write_csv(sigMarkers,"analysis_outs/cluster_markers_full.csv")
```

## [Save RDS]{.underline}

```{r}
saveRDS(dataSubQC, "C:/Users/Eric/My Drive/Lab/datasets/EYW/CTR03_10x_240801/seurat_outs/CTR03_full.rds")
```
