---
title: "DE MAST analysis"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(patchwork)
library(MAST)
library(future)
library(furrr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
source("functions/scRNA_seq_analysis_functions.R")
theme_set(theme_Publication())
```

## [Import data]{.underline}

```{r}
data <- readRDS("C:/Users/Eric/Documents/datasets/EYW/SIG04_10x_240816/seurat_outs/SIG04_postQC_seurat_all.rds")
# subset to singlets only
data <- subset(data, subset = oBC_classification == "singlet")
```


## [Overview]{.underline}

```{r}
data@meta.data %>%
  group_by(oBC_feature_call) %>%
  summarise(n = n()) 

data@meta.data %>%
  filter(gem_group == "lane1") %>%
  group_by(oBC_feature_call) %>%
  summarise(n = n()) 

data@meta.data %>%
  filter(gem_group == "lane2") %>%
  group_by(oBC_feature_call) %>%
  summarise(n = n()) 
```


```{r}
p1 <- data@meta.data %>%
  group_by(oBC_feature_call) %>%
  summarise(n = n()) %>%
    ggplot(aes(x="full",y=n)) +
      geom_violin() +
      geom_boxplot(width=0.1) +
      xlab("") +
      ylab("cell #")

p2 <- data@meta.data %>%
  group_by(gem_group, oBC_feature_call) %>%
  summarise(n = n()) %>%
    ggplot(aes(x="",y=n, fill = gem_group)) +
      geom_violin() +
      geom_boxplot(width=0.1, position = position_dodge(0.9)) +
      scale_fill_brewer(palette="Dark2")+
      xlab("") +
      ylab("cell #")

p1+p2
```


## DEG 

I want to regress out the effects of more or less virus. Let's see if I can use the p139 mapped counts from the GEX data or if I should use the oBC counts

```{r}
# create metadata columns of scaled viral counts for use as covariate
data$p139_counts = FetchData(data, vars = "p139-T7oBC5p-MS2", layer = "scale.data")
```

```{r, fig.height=5, fig.width=10}
library(ggpointdensity)

data@meta.data %>%
  ggplot(aes(x=p139_counts, y=oBC_log1p_total_counts)) +
    geom_pointdensity() +
    scale_color_viridis_c() +
    facet_wrap(~gem_group) +
    theme(aspect.ratio = 1)
```

I feel like the p139 transcript counts are a bit too sparse to be accurate. Lets use the oBC counts

### DEG for full dataset

Perform MAST DEG analysis accounting for gem_group and viral counts.
```{r}
# helper function for calculating DEGs
deg_MAST <- function(data,x,y){
  DefaultAssay(data) <- "RNA"
  Idents(data) <- "oBC_feature_call"
  # perform DEG analysis using seurat wrapper
  test <- FindMarkers(data,
              slot = "data", #MAST uses log normalized counts as input
              min.pct = 0.05,
              logfc.threshold=0.1,
              ident.1 = x, ident.2 = y,
              test.use = "MAST",
              latent.vars = c("gem_group","oBC_log1p_total_counts"))
  
  # make tibble tidy
  test <- test %>%
    as_tibble(rownames="genes") %>%
    mutate(comp_1 = x,
           comp_2 = y)
  
  return(test)
}

# get list of ligands (comp_1)
ligands <- data$oBC_feature_call %>% unique()
ligands <- ligands[grep("p129",ligands, invert = T)]
```

```{r}
# set options for parallel analysis
plan(multisession, workers=parallelly::availableCores())
```


```{r}
options(future.globals.maxSize= 10000*1024^2)
# degFull <- tibble()
# for(ligand in ligands ){
#   temp <- deg_MAST(data,ligand,"p129")
#   degFull <- bind_rows(degFull,temp)
# }

degFull <- future_map_dfr(c("IL4","IL4-pool","IL6","IL6-pool","IFNG","IFNG-pool","IL12","IL12-pool","IFNA","IFNA-pool"), ~ deg_MAST(data, .x, "p129"), seed=T)
```

```{r}
write_csv(degFull, "DEG_MAST_full.csv")
```

```{r}
degFullSig <- degFull %>%
  filter(p_val_adj < 0.1) %>%
  group_by(comp_1) %>%
  head(n=200)
```

```{r}
dataSubPool <- subset(data, subset = oBC_feature_call %in% c("IL4","IL4-pool","IL6","IL6-pool","IFNG","IFNG-pool","IL12","IL12-pool","IFNA","IFNA-pool"))
dataSubPool <- ScaleData(dataSubPool, features = rownames(dataSubPool))


```

```{r, fig.height=20, fig.width=10}
Idents(dataSubPool) <- "oBC_feature_call"
DoHeatmap(dataSubPool, features = degFullSig$genes %>% unique())
```
```{r}
mat <- GetAssayData(dataSubPool,assay = "RNA", layer = "scale.data")
rownames(mat) <- rownames(dataSubPool@assays$RNA@layers$scale.data)
mat <- mat[rownames(mat)[rownames(mat) %in% unique(degFullSig)],]
```

```{r}
library(ComplexHeatmap)

pheatmap(mat = mat)
```


### DEG for each lane2 only

Perform MAST DEG analysis accounting for gem_group and viral counts.
```{r}
dataLane2 <- subset(data, subset = gem_group == "lane2")

degLane2 <- tibble()
for(ligand in ligands){
  temp <- deg_MAST(dataSub,ligand,"p129")
  degLane2 <- bind_rows(degLane2,temp)
}
```

```{r}
write_csv(degFull, "DEG_MAST_lane2_sorted.csv")
```
























