---
title: "SIG04 DEG visualizations"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(ggplot2)
library(patchwork)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
source("functions/scRNA_seq_analysis_functions.R")
theme_set(theme_Publication())
```

## [Import data]{.underline}

```{r}
# import seurat object and subset to singlets
data <- readRDS("C:/Users/Eric/Documents/datasets/EYW/SIG04_10x_240816/seurat_outs/SIG04_postqc_seurat.rds")
data <- subset(data, subset = oBC_classification == "singlet")
```

```{r}
meta <- read_csv("processing_outs/plate_meta_SIG04.csv")
```

## [Barcode Expression Heatmap]{.underline}

```{r}
metadata <- data@meta.data %>%
  as_tibble(rownames = "cell_bc")

# extract CLR normalized counts and add feature call metadata
normCounts <- GetAssayData(data, assay = "oBC", layer = "data") %>%
  as_tibble(rownames = "ligand") %>%
  pivot_longer(-ligand, names_to = "cell_bc", values_to = "expression") %>%
  left_join(metadata, by="cell_bc")

# summarize mean expression
summaryCounts <- normCounts %>%
  group_by(oBC_feature_call, ligand) %>%
  summarise(mean_exp = mean(expression))
```

```{r, fig.height=14, fig.width=16}
summaryCounts %>%
  filter(!grepl("-pool",oBC_feature_call)) %>%
  filter(!grepl("-pool",ligand)) %>%
  ggplot(aes(x=oBC_feature_call,y=ligand,fill=mean_exp)) +
    geom_tile() +
    scale_fill_viridis_c() +
    ggtitle("oBC Mean CLR normalized counts in Feature Calls") +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r, fig.height=14, fig.width=16}
summaryCounts %>%
  ggplot(aes(x=oBC_feature_call,y=ligand,fill=mean_exp)) +
    geom_tile() +
    scale_fill_viridis_c() +
    ggtitle("oBC Mean CLR normalized counts in Feature Calls") +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## [Barcode MFI correlations]{.underline}

First, lets see if barcode counts correlate with mBaoJin MFI by flow. I am mainly concerned with whether there is some independent assortment of barcode containing virus and ligand displaying virus.

We'll filter to only include cells from lane 1 because lane 2 was sorted so it's not really representative of raw MFI -> barcode correlations

```{r}
metadata <- data@meta.data %>%
  as_tibble(rownames = "cell_bc")

# extract CLR normalized counts and add feature call metadata
normCounts <- GetAssayData(data, assay = "oBC", layer = "data") %>%
  as_tibble(rownames = "ligand") %>%
  pivot_longer(-ligand, names_to = "cell_bc", values_to = "expression") %>%
  left_join(metadata, by="cell_bc")

# summarize mean expression
# filter by lane 1
summaryCounts <- normCounts %>%
  filter(gem_group == "lane1") %>%
  group_by(oBC_feature_call, ligand) %>%
  summarise(mean_exp = mean(expression),
            mean_log1p_umi = mean(oBC_log1p_total_counts))

# join summary counts from above with MFI
# filter to only include matching feature_call ligand pairs
summaryCountsCalled <- summaryCounts %>%
  left_join(meta, by="oBC_feature_call") %>%
  filter(oBC_feature_call == ligand)
```

```{r, fig.height=5, fig.width=10}
p1 <- summaryCountsCalled %>%
  ggplot(aes(x=mBaojin_MFI_log10,y=mean_exp)) +
    geom_point() +
    ylab("oBC mean norm expression\nof respective barcode") +
    theme(aspect.ratio = 1)

p2 <- summaryCountsCalled %>%
  ggplot(aes(x=mBaojin_MFI_log10,y=mean_exp)) +
    geom_point() +
    geom_text_repel(aes(label = oBC_feature_call)) +
    ylab("oBC mean norm expression\nof respective barcode") +
    theme(aspect.ratio = 1)

p1+p2
```

There's a good trend with flow MFI and barcode detection. IL12 is kind of an outlier which probably explains why I didn't see anything in the perturbation.

```{r}
# summarize mean expression
# filter by lane 1
summaryCounts <- normCounts %>%
  group_by(gem_group, oBC_feature_call, ligand) %>%
  summarise(mean_exp = mean(expression))

# join summary counts from above with MFI
# filter to only include matching feature_call ligand pairs
summaryCountsCalled <- summaryCounts %>%
  left_join(meta, by="oBC_feature_call") %>%
  filter(oBC_feature_call == ligand)

summaryCountsCalled %>%
  ggplot(aes(x = gem_group, y = mean_exp, fill = gem_group)) +
    geom_line(aes(group = oBC_feature_call)) +
    geom_boxplot() +
    geom_point() +
    ylab("oBC mean norm expression\nof respective barcode") +
    theme(aspect.ratio = 1)
```

As expected lane 2, which is flow sorted for mBaoJin+ cells has a tighter distribution especially towards the bottom end. The distributions are a bit shifted because the lane 2 amplification of oBC was worse for some reason. 















