---
title: "Clustering Detailed visualizations"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(ggplot2)
library(patchwork)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
source("functions/scRNA_seq_analysis_functions.R")
theme_set(theme_Publication())
```

## [Import data]{.underline}

```{r}
dataLog1p <- read_csv("analysis_outs/mean_aggregated_normUMI_log1p_lane2.csv") %>%
  pivot_longer(-oBC_feature_call, names_to = "genes", values_to = "expression")
dataZscore <- read_csv("analysis_outs/mean_aggregated_normUMI_Zscore_lane2.csv") %>%
  pivot_longer(-oBC_feature_call, names_to = "genes", values_to = "expression")
```

## [Scatterplot Visualizations]{.underline}

### Believable Clusters

Let's take a look at what the scatterplots look like for the more believable
structure in the dataset

```{r}
scatterViz <- function(data,x,y,label=FALSE,label_x=NA,label_y=NA){
  data <- filter(data, oBC_feature_call %in% c(x,y)) %>%
    pivot_wider(id_cols = genes, names_from = oBC_feature_call, values_from = expression)
  
  if(label == F){
     p1 <- ggplot(data, aes_string(x=x, y=y)) +
      geom_point() +
      theme(aspect.ratio = 1)
  }
 
  if(label == T){
    # make column for labels
    data$label <- ifelse(data[[x]] > label_x | data[[x]] < -label_x | data[[y]] > label_y | data[[y]] < -label_y,
                     as.character(data$genes), NA)
    
    p1 <- ggplot(data, aes_string(x=x, y=y)) +
      geom_point() +
      geom_text_repel(aes(label = label)) +
      theme(aspect.ratio = 1)
  } 
  
  return(p1)
}
```

```{r, fig.height=10, fig.width=15}
scatterViz(dataZscore, "IFNG","IL27", label = T, label_x = 1, label_y = 1) +
  scatterViz(dataZscore, "IFNA","IFNB1", label = T, label_x = 2, label_y = 2) +
  scatterViz(dataZscore, "IFNA","IFNK", label = T, label_x = 2, label_y = 2) +
  scatterViz(dataZscore, "IL6","IL21", label = T, label_x = 1, label_y = 1) +
  scatterViz(dataZscore, "IL6","IL11", label = T, label_x = 1.5, label_y = 0.2)
```

```{r, fig.height=10, fig.width=15}
scatterViz(dataLog1p, "IFNG","p129") +
  scatterViz(dataLog1p, "IFNA","p129") +
  scatterViz(dataLog1p, "IL6","p129") +
  scatterViz(dataLog1p, "IL21","p129") +
  scatterViz(dataLog1p, "IL11","p129") +
  scatterViz(dataLog1p, "IL7","p129")
```

### GDF2/CCL25/CXCL11

```{r, fig.width=15, fig.height=5}
scatterViz(dataZscore, "GDF2","CCL25", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "GDF2","CXCL11", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "CXCL11","CCL25", label = T, label_x = 0.2, label_y = 0.2)
```

### TNF/OSM/TSLP

```{r, fig.width=15, fig.height=5}
scatterViz(dataZscore, "TNF","OSM", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TNF","TSLP", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "OSM","TSLP", label = T, label_x = 0.2, label_y = 0.2)
```

### ADIPOQ Cluster

```{r, fig.width=15, fig.height=5}
scatterViz(dataZscore, "ADIPOQ","TGFB3", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "ADIPOQ","IL15", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "ADIPOQ","IL1A", label = T, label_x = 0.2, label_y = 0.2)
```

### TGFBs

```{r, fig.width=15, fig.height=5}
scatterViz(dataZscore, "TGFB1","TGFB3", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TGFB2","TGFB3", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TGFB1","TGFB2", label = T, label_x = 0.2, label_y = 0.2)
```

### Cluster 2

We can see that IL1A, IL15, ADIPOQ, and TGFB3 form a small cluster while MIF,
TNF, IL9, OSM, and TSLP form a different cluster. There is quite a bit of
correlation between these two though.

```{r, fig.width=15, fig.height=10}
scatterViz(dataZscore, "TNF","OSM", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TSLP","OSM", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TNF","TSLP", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "IL1A","IL15", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "ADIPOQ","IL15", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TGFB3","IL15", label = T, label_x = 0.2, label_y = 0.2) +
  plot_layout(ncol=3)
```

```{r, fig.height=5, fig.width=10}
scatterViz(dataZscore, "CXCL9","CXCL10", label = T, label_x = 0.5, label_y = 0.5) +
  scatterViz(dataZscore, "CXCL9","IFNA", label = T, label_x = 0.5, label_y = 0.5)
```

```{r}
scatterViz(dataZscore, "IL12","p129", label = T, label_x = 0.2, label_y = 0.2)
```

```{r, fig.width=15, fig.height=5}
scatterViz(dataZscore, "TGFB1","TGFB2", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TGFB1","TGFB3", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "TGFB2","TGFB3", label = T, label_x = 0.2, label_y = 0.2)
```

```{r, fig.width=15, fig.height=10}
scatterViz(dataZscore, "GDF2","CCL25", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "CXCL11","CCL25", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "GDF2","CXCL11", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "GDF2","TNF", label = T, label_x = 0.2, label_y = 0.2) +
  plot_layout(ncol=3)
```

```{r, fig.height=10, fig.width=20}
scatterViz(dataZscore, "TNFSF15","IL16", label = T, label_x = 0.1, label_y = 0.1) +
  scatterViz(dataZscore, "WNT1","IL16", label = T, label_x = 0.1, label_y = 0.1)
```

```{r, fig.height=10, fig.wdith=15}
scatterViz(dataZscore, "IL17F","IL17A", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "IL25","IL17A", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "IL17B","IL17A", label = T, label_x = 0.2, label_y = 0.2) +
  scatterViz(dataZscore, "WNT1","IL17A", label = T, label_x = 0.2, label_y = 0.2)
```
